Well-quasi-orderings on word languages

### 2025-08-06 src/prefixes.tex
removing todos + polish
more notes

Let us now go further and fully characterize languages $L$ such that the
prefix relation is well-quasi-ordered, without any restriction on the
decidability of $L$ itself. Let us remark that finite unions of \kl{chains} are
always \kl{well-quasi-ordered} by the \kl{prefix relation} because they lack
infinite \kl{antichains} by definition. The following theorem states that this
is the only possible reason for a language $L$ to be \kl{well-quasi-ordered} by
the \kl{prefix relation}.

For the proof of the following theorem, we will use special notations to
describe the \intro{upwards closure} of a set $S$ for a relation $\preceq$,
which is defined as $\upset[\preceq]{S} \defined \setof{y \in \Sigma^*}{
\exists x \in S. x \preceq y}$. Anticipating the use of the symmetric notion of
\intro{downwards closure}, let us introduce it as follows: $\dwset[\preceq]{S} \defined \setof{y
\in \Sigma^*}{ \exists x \in S. y \preceq x}$. Abusing notations, we will
write $\upset{w}$ and $\dwset{w}$ for the upwards and downwards closure of a
single element $w$, omitting the ordering relation when it is clear from the
context. A set $S$ is called \intro{downwards closed} if $\dwset{S} = S$.

@@

    Conversely, assume that $L$ is \kl{well-quasi-ordered} by the \kl{prefix
    relation}. Let us define $S$ the set of words $w$ such that there exists
    two words $wu$ and $wv$ both in $L$ that are not comparable for the
    \kl{prefix relation}. Assume by contradiction that $S$ is infinite.
    Then, $S$ equipped with the \kl{prefix relation} is an infinite
    tree with finite branching, and therefore contains an infinite
    branch, which is by definition an \kl{antichain branch} for $L$.
    This contradicts the assumption that $L$ is \kl{well-quasi-ordered}.
    Now, consider $w$ be a maximal element for the \kl{prefix ordering}
    in $S$. By definition, all words in $L$ that contain $w$ as a prefix
    must be comparable for the \kl{prefix relation}. Therefore,
    $(\upset[\prefleq]{w}) \cap L$ is a \kl{chain} for the \kl{prefix relation}.
    In particular, letting $S_{\max}$ be the set of all maximal elements
    of $S$,
    we conclude that 
    \begin{equation*}
        L \subseteq S \cup \bigcup_{w \in S_{\max}} (\upset[\prefleq]{w}) \cap L
        \quad .
    \end{equation*}
    Hence, that $L$ is finite union of \kl{chains}.
\end{proof}

@@

\begin{corollary}
    \label{prefixes-ordinal-invariants:cor}
    Let $L \subseteq \Sigma^*$ be a language that is
    \kl{well-quasi-ordered} by the \kl{prefix relation}. Then,
    \kl{maximal order type} of $L$ strictly smaller than $\omega^2$,
    the \kl{ordinal height} of $L$ is at most $\omega$, and
    its \kl{ordinal width} is finite.
\end{corollary}

\todo[inline]{
    We actually can \emph{compute} the ordinal invariants
    in the regular case.
}

>>>>

Let us now go further and fully characterize languages $L$ such that the
prefix relation is well-quasi-ordered, without any restriction on the
decidability of $L$ itself.
For the proof of the following theorem, we will use special notations to
describe the \intro{upwards closure} of a set $S$ for a relation $\preceq$,
which is defined as $\upset[\preceq]{S} \defined \setof{y \in \Sigma^*}{
\exists x \in S. x \preceq y}$. Anticipating the use of the symmetric notion of
\intro{downwards closure}, let us introduce it as follows: $\dwset[\preceq]{S} \defined \setof{y
\in \Sigma^*}{ \exists x \in S. y \preceq x}$. Abusing notations, we will
write $\upset{w}$ and $\dwset{w}$ for the upwards and downwards closure of a
single element $w$, omitting the ordering relation when it is clear from the
context. A set $S$ is called \intro{downwards closed} if $\dwset{S} = S$.

@@

    Conversely, assume that $L$ is \kl{well-quasi-ordered} by the \kl{prefix
    relation}. Let us define $S$ the set of words $w$ such that there exists
    two words $wu$ and $wv$ both in $L$ that are not comparable for the
    \kl{prefix relation}. Assume by contradiction that $S$ is infinite.
    Then, $S$ equipped with the \kl{prefix relation} is an infinite
    tree with finite branching, and therefore contains an infinite
    branch, which is by definition an \kl{antichain branch} for $L$.
    This contradicts the assumption that $L$ is \kl{well-quasi-ordered}.
    Now, consider $w$ be a maximal element for the \kl{prefix ordering}
    in $S$. By definition, all words in $L$ that contain $w$ as a prefix
    must be comparable for the \kl{prefix relation}. Therefore,
    $(\upset[\prefleq]{w}) \cap L$ is a \kl{chain} for the \kl{prefix relation}.
    In particular, letting $S_{\max}$ be the set of all maximal elements
    of $S$,
    we conclude that 
    \begin{equation*}
        L \subseteq S \cup \bigcup_{w \in S_{\max}} (\upset[\prefleq]{w}) \cap L
        \quad .
    \end{equation*}
    \todo[inline]{Because $S$ only contains words with two incomparable suffixes in $L$, I think this is wrong - consider e.g. $L \subseteq a^*$, where $S$ and $S_{max}$ are necessarily empty}
    Hence, that $L$ is finite union of \kl{chains}.
\end{proof}

@@

\begin{corollary}
    \label{prefixes-ordinal-invariants:cor}
    Let $L \subseteq \Sigma^*$ be a language that is
    \kl{well-quasi-ordered} by the \kl{prefix relation}. Then,
    \kl{maximal order type} of $L$ strictly smaller than $\omega^2$,
    the \kl{ordinal height} of $L$ is at most $\omega$, and
    its \kl{ordinal width} is finite. Furthermore, these bounds 
    are tight.
\end{corollary}
\begin{proof}
    The upper bounds follow from the fact that $L$ is a finite union of
    \kl{chains}. The tightness can be obtained by considering the languages
    $L_k \defined \bigcup_{i=0}^{k-1} a^i b^*$ for $k \in \Nat$, which 
    are \kl{well-quasi-ordered} by the \kl{prefix relation} (as they are
    finite unions of \kl{chains}), and satisfy
    that $\oWidth{L_k} = k$, $\oHeight{L_k} = \omega$,
    and therefore $\oType{L_k} = k \cdot \omega$.
\end{proof}
